The Economics of Wildfire Aviation

### 2023-10-16 manuscript/manuscript.qmd
updated ms in response to reviewer comments

We examine two economic aspects of aviation used for wildland firefighting in the United States. First, we analyze the procurement of aviation resources by federal agencies, and how the structure of contracts influences suppression costs. Second, we highlight the lack of empirical evidence on the effectiveness of aviation, and provide a research design to quantify effectiveness and the benefits of aviation use. Our analysis can inform tactical operations on individual wildfire incidents and help federal agencies develop season-long resource utilization strategies.
[^ack]

[^ack]: Acknowledgements: We acknowledge support from the United States Forest Service (agreement #21-CS-11221636-151). We are thankful for the research assistance from Kyra Adamson.

@@

<!-- aviation expense -->

@@

Wildfire response is a coordinated effort between federal, state, and local land management agencies. Firefighting resources include hand crews, engines, dozers and aircraft. These resources are owned by different agencies but will coordinate under unified command when incidents become large and complex. When multiple fires are burning, resources are allocated according to need and agency official's expectation of benefit on specific fires [@bayham2020].

Aviation resources have been used in wildfire response since WWII [@pyne1982] when military aircraft were retrofit with equipment that would allow them to carry and drop water. Over the years, government agencies have acquired, retrofit, and maintained a fleet of airplanes and helicopters. These aircraft have been used to drop water and retardant on or in advance of the fire, deploy smoke jumpers, to gather information about the extent and intensity of the fire, and to survey damage. While aviation can provide valuable services, they are one of the most expensive resources on an active fire.

@@

Until 2004, federal agencies owned and operated a fleet of aircraft bearing the cost of maintenance. Since 2004, federal agencies have contracted some aviation resources owned by private entities [@thompson2012]. The contacts specify a number of safety requirements, but also a daily availability price and an hourly flight price. The two forms of contract are Exclusive Use (EXU) and Call-When-Needed (CWN). Exclusive use contracts mandate that an aircraft is exclusively available to the contracting entity (e.g., US Forest Service) for a specified number of days at a daily cost. Call-when-needed contracts are more flexible and can be activated on short notice [@thompson2012]. However, CWN contacts generally specify a daily availability rate and hourly flight rate that is higher than EXU contracts. The question is: how many of each contract should the USFS sign at the start of each fire season?[^1] The answer depends on expected use patterns of the aviation resources.

[^1]: The Rand Corporation was asked to answer this question in 2012. They concluded that the USFS should make more and better use of "scooper" aircraft that are capable of refilling water tanks by flying over water bodies. The results of this report were largely dismissed by leaders in the firefighting community for making heroic assumptions about the ability of scoopers [@keating2012].

@@

@fig-aviation_use illustrates the variability and uncertainty of initial attack fires (bottom panel) and large fires being actively suppressed (extended attack - top panel) throughout the season. According to the 2022 USFS Schedule of Items, most large airtankers under exclusive use contracts started between March and May and were scheduled to end in September, October, or November.[^2] While the 18 large and very large airtankers under contract in 2022 span the dates when fire activity is expected to be high, they are not activated in a pattern that follows the number of IA and EA fires in @fig-aviation_use. The trends in @fig-aviation_use suggest that given current usage patterns, the USFS could delay activating some LATs and VLATs until later in June. Call When Needed contracts could be used to handle surge capacity and unexpected increases in fire activity.

[^2]: The only publicly accessible document we could locate was from June 24, 2022 https://www.fs.usda.gov/Internet/FSE_DOCUMENTS/fseprd1041508.pdf

@@

Cost benefit analysis requires the accounting of the costs and benefits of an intervention over a relevant time period [@boardman2018]. The direct cost of using aviation to fight fire is recorded because the government is responsible for payment.[^3] The benefits of aviation are harder to quantify because they involve the avoided costs and losses of fires that were prevented from growing and damaging valued assets. If aviation resources reduce the probability that a fire spreads and damages property, we do not observe the counterfactual - what would have happened in the absence of aviation use. This section describes how aviation resources are currently used to fight fire, discusses the conceptual challenges to measuring benefits, and proposes data driven approaches to measuring the benefits of aviation given currently available data.

[^3]: The indirect costs of aviation include the risk to pilots that fly in dangerous conditions around challenging terrain. @butler2015 document that 78 of the 298 (26%) recorded wildland firefighting fatalities were attributable to aviation incidents between 2000 and 2013.

Aviation resources such as large airtankers (LAT) are believed to be most effective during the initial attack phase (the first burning period which is usually from ignition to 10:00 AM the following day) [@calkin2014a]. The logic is simple, aggressively attack the fire while it is small and it will not grow large. Most fires (more than 90%) are contained during initial attack [@holmes2008]. However, the fires that do escape initial containment because of weather or lack of sufficient IA effort can grow large and threaten values at risk.[^4]

[^4]: Values at risk is an umbrella term that captures assets that society values, including private property, infrastructure, cultural heritage sites, endangered species habitat.

@@

@fig-example_drop illustrates the empirical strategy on a retardant drop on the Terek fire that burned in Wyoming in 2018. The fire is progressing south and retardant is dropped in advance of the fire. The drop is 529 meters long. The north side of the drop (A) is the pre-treatment region while the south side (B) is the post-treatment region. The control region is east of the retardant line with the north and south similarly defined as the pre (C) and post period (D)A. In this example, the fire does not progress south of the drop into region B, but it does progress into region D (post-period control area), providing some evidence that the drop prevented the spread of the fire to the south.

@@

$$ y_{it} = \alpha + \beta_1 Treated_i + \beta_2 Post_t + \beta_3 (Treated_i \cdot Post_t) + \gamma X_{it} + \varepsilon_{it} , $$ where $y_{it}$ is the outcome variable for unit $i$ at time $t$. The outcome could be a binary indicator for whether an area burned or a measure of intensity like Fire Radiance Power.[^5] $Treated_i$ is a binary variable indicating whether unit $i$ is in the treatment group (region A and B in @fig-example_drop). $Post_t$ is a binary variable indicating whether the observation is from the post-treatment period (region B and D in @fig-example_drop). $X_{it}$ is a vector of vegetation measures, topography, and weather, and $\varepsilon_{it}$ is the error term. The coefficient of interest is $\beta_3$, which represents the treatment effect. It measures the difference in the change in fire outcomes between the treatment and control groups before and after the treatment.

[^5]: Fire Radiance Power is a quantitative measure of energy detected by satellites observing fire activity. These data are generated and shared by *NASA's Fire Information for Resource Management System (FIRMS) (https://earthdata.nasa.gov/firms), part of NASA's Earth Observing System Data and Information System (EOSDIS).*

@@

This framework is simple but powerful. First, it produces a causal estimate of the average effectiveness of large airtanker drops toward stopping the spread of fire. Second, we can easily investigate heterogeneity of the treatment effect by interacting topography, vegetation, and weather with $(Treated_i \cdot Post_t)$. These results can help inform strategic operations, highlighting when and where large airtanker drops are most effective. The results also serve as a crucial step in estimating the benefits - avoided damage - of LAT drops. For example, at time $t$, assume there is some risk to assets denominated by $\rho(e_0)*V$ where $\rho(e_0)$ is a probability of loss as a function of effort, $e_0$, and $V$ is the dollar value of the assets at risk.[^6] If a LAT drop represents effort $e_1 > e_0$, then the benefit of the drop is $V*\left[\rho(e_0) - \rho(e_1) \right]$. This benefit is comparable to the known cost of the flight based on the hourly cost and cost of retardant.

[^6]: Assets like homes generally have observable values, whereas cultural heritage sites may not. Moreover, many object to the strict use of monetary value of assets that may imply the benefits of protecting high-value homes are higher than protecting low-value homes. The principles described here accommodate alternative definitions of values at risk that may include subjective values.

@@

<!-- high level connecting wildfire to climate change; mitigation is not enough -->

A changing climate is driving longer and more intense fire seasons. There is an urgent need to reduce wildfire risk in fire-prone regions of the US. Yet, the primary method for reducing risk at the landscape level involves the costly removal of combustible fuels. Despite significant progress, the number of acres in need of treatment is very large and not expected to be treated over the next decade [@thompson2023]. Wildfire suppression and response currently does and will continue to play an important role in protecting people, property, and other valued assets from wildfire. As federal firefighting agencies seek to improve efficient use of resources, the procurement and use of aviation resources presents several opportunities.

The current structure of procuring aviation resources involves exclusive use and call-when-needed contracts. Federal agencies should strive to align their use of aviation with operational demands. However, risk aversion may lead agency administrators to over contract for aviation resources to avoid public and political consequences that arise after destructive wildfire events.

@@

![Trends of Extended Attack (top) and Initial Attack (middle) fires from 2007 to 2021. The points represent the median and the lines represent the interquartile range (25th to 75th percentile). The bottom panel represents the number of LATs and VLATs available by day under exclusive use contract in 2022. The initial and extended attack fire counts are compiled from historical daily situation reports posted at https://www.nifc.gov/nicc-files/sitreprt.pdf and collected using code from [@nguyen2023], and the number of LATs and VLATs contracted are from the June 24, 2022 Schedule of Items posted at https://www.fs.usda.gov/Internet/FSE_DOCUMENTS/fseprd1041508.pdf](../figures/ia_ea_fires.png){#fig-aviation_use}

>>>>

We examine two economic aspects of aviation used for wildland firefighting in the United States. First, we analyze the procurement of aviation resources by federal agencies, and how the structure of contracts influences suppression costs. Second, we highlight the lack of empirical evidence on the effectiveness of aviation, and provide a research design to quantify effectiveness and the benefits of aviation use. Our analysis can inform tactical operations on individual wildfire incidents and help federal agencies develop season-long resource utilization strategies. [^1]

[^1]: Acknowledgements: We acknowledge support from the United States Forest Service (agreement #21-CS-11221636-151). We are thankful for the research assistance from Kyra Adamson.

@@

@@

Wildfire response is a coordinated effort between federal, state, and local land management agencies. Firefighting resources include hand crews, engines, dozers and aircraft. These resources are owned by different agencies but will coordinate under unified command when incidents become large and complex. When multiple fires are burning, resources are allocated according to need and agency official's expectation of benefit on specific fires [@bayham2020]. Aircraft are one of the most visible firefighting resources and are used to drop water and retardant on or in advance of the fire, deploy smoke jumpers, to gather information about the extent and intensity of the fire, and to survey damage. While aviation can provide valuable services, they are one of the most expensive resources on an active fire.

@@

Aircraft have been used in wildfire response since WWII [@pyne1982] when military aircraft were retrofit with equipment that would allow them to carry and drop water. Over the years, government agencies have acquired, retrofit, and maintained a fleet of airplanes and helicopters, some of which were owned and others leased from private entities. While the fleet of firefighting aircraft is diverse including various sizes of helicopter, small airplanes, and modified commercial jets, much of the federal aviation fleet is now leased from private entities. The reasons for leasing aircraft include the high cost of ownership, flexibility associated with variation in fire activity throughout the year, the ability use modern aircraft without having to frequently cycle inventory [@thompson2012].

Aviation contracts are complex legal documents, but generally include a number of safety requirements, a daily availability price, and an hourly flight price. The two forms of contract are Exclusive Use (EXU) and Call-When-Needed (CWN). Exclusive use contracts mandate that an aircraft is exclusively available to the contracting entity (e.g., US Forest Service) for a specified number of days at a daily cost. Call-when-needed contracts are more flexible and can be activated on short notice [@thompson2012]. However, CWN contacts generally specify a daily availability rate and hourly flight rate that is higher than EXU contracts. The question is: how many of each contract should the USFS sign at the start of each fire season?[^2] The answer depends on expected use patterns of the aviation resources. We focus on this question in the context of large and expensive aircraft known as Large Airtankers (LAT) and Very Large Airtankers (VLAT) classified based on their water and retardant capacity.

[^2]: While the agency may also choose to own and maintain their own aircraft, we only consider the choice between existing contract types because the USFS has shown a reluctance to own and maintain aircraft in recent years.

@@

@fig-aviation_use illustrates the variability and uncertainty of initial attack fires (bottom panel) and large fires being actively suppressed (extended attack - top panel) throughout the season. According to the 2022 USFS Schedule of Items, most large airtankers under exclusive use contracts started between March and May and were scheduled to end in September, October, or November.[^3] While the 18 large and very large airtankers under contract in 2022 span the dates when fire activity is expected to be high, they are not activated in a pattern that follows the number of IA and EA fires in @fig-aviation_use. The trends in @fig-aviation_use suggest that given current usage patterns, the USFS could delay activating some Large Airtankers (LAT) and Very Large Airtankers (VLAT) until later in June. Call When Needed contracts could be used to handle surge capacity and unexpected increases in fire activity.

[^3]: The only publicly accessible document we could locate was from June 24, 2022 https://www.fs.usda.gov/Internet/FSE_DOCUMENTS/fseprd1041508.pdf

@@

Cost benefit analysis requires the accounting of the costs and benefits of an intervention over a relevant time period [@boardman2018]. The direct cost of using aviation to fight fire is recorded because the government is responsible for payment.[^4] The benefits of aviation are harder to quantify because they involve the avoided costs and losses of fires that were prevented from growing and damaging valued assets. If aviation resources reduce the probability that a fire spreads and damages property, we do not observe the counterfactual - what would have happened in the absence of aviation use. This section describes how aviation resources are currently used to fight fire, discusses the conceptual challenges to measuring benefits, and proposes data driven approaches to measuring the benefits of aviation given currently available data.

[^4]: The indirect costs of aviation include the risk to pilots that fly in dangerous conditions around challenging terrain. @butler2015 document that 78 of the 298 (26%) recorded wildland firefighting fatalities were attributable to aviation incidents between 2000 and 2013.

Aviation resources such as large airtankers (LAT) are believed to be most effective during the initial attack phase (the first burning period which is usually from ignition to 10:00 AM the following day) [@calkin2014a]. The logic is simple, aggressively attack the fire while it is small and it will not grow large. Most fires (more than 90%) are contained during initial attack [@holmes2008]. However, the fires that do escape initial containment because of weather or lack of sufficient IA effort can grow large and threaten values at risk.[^5]

[^5]: Values at risk is an umbrella term that captures assets that society values, including private property, infrastructure, cultural heritage sites, endangered species habitat.

@@

@fig-example_drop illustrates the empirical strategy on a retardant drop on the Terek fire that burned in Wyoming in 2018. The fire is progressing south and retardant is dropped in advance of the fire. The drop is 529 meters long. The north side of the drop (A) is the pre-treatment region while the south side (B) is the post-treatment region. The control region is east of the retardant line with the north and south similarly defined as the pre (C) and post period (D). In this example, the fire does not progress south of the drop into region B, but it does progress into region D (post-period control area), providing some evidence that the drop prevented the spread of the fire to the south.

@@

$$ y_{it} = \alpha + \beta_1 Treated_i + \beta_2 Post_t + \beta_3 (Treated_i \cdot Post_t) + \gamma X_{it} + \varepsilon_{it} , $$ where $y_{it}$ is the outcome variable for unit $i$ at time $t$. The outcome could be a binary indicator for whether an area burned or a measure of intensity like Fire Radiance Power.[^6] $Treated_i$ is a binary variable indicating whether unit $i$ is in the treatment group (region A and B in @fig-example_drop). $Post_t$ is a binary variable indicating whether the observation is from the post-treatment period (region B and D in @fig-example_drop). $X_{it}$ is a vector of vegetation measures, topography, and weather, and $\varepsilon_{it}$ is the error term. The coefficient of interest is $\beta_3$, which represents the treatment effect. It measures the difference in the change in fire outcomes between the treatment and control groups before and after the treatment.

[^6]: Fire Radiance Power is a quantitative measure of energy detected by satellites observing fire activity. These data are generated and shared by *NASA's Fire Information for Resource Management System (FIRMS) (https://earthdata.nasa.gov/firms), part of NASA's Earth Observing System Data and Information System (EOSDIS).*

@@

This framework is simple but powerful. First, it produces a causal estimate of the average effectiveness of large airtanker drops toward stopping the spread of fire. Second, we can easily investigate heterogeneity of the treatment effect by interacting topography, vegetation, and weather with $(Treated_i \cdot Post_t)$. These results can help inform strategic operations, highlighting when and where large airtanker drops are most effective. The results also serve as a crucial step in estimating the benefits - avoided damage - of LAT drops. For example, at time $t$, assume there is some risk to assets denominated by $\rho(e_0)*V$ where $\rho(e_0)$ is a probability of loss as a function of effort, $e_0$, and $V$ is the dollar value of the assets at risk.[^7] If a LAT drop represents effort $e_1 > e_0$, then the benefit of the drop is $V*\left[\rho(e_0) - \rho(e_1) \right]$. This benefit is comparable to the known cost of the flight based on the hourly cost and cost of retardant.

[^7]: Assets like homes generally have observable values, whereas cultural heritage sites may not. Moreover, many object to the strict use of monetary value of assets that may imply the benefits of protecting high-value homes are higher than protecting low-value homes. The principles described here accommodate alternative definitions of values at risk that may include subjective values.

@@

A changing climate is driving longer and more intense fire seasons. There is an urgent need to reduce wildfire risk in fire-prone regions of the US. Yet, the primary method for reducing risk at the landscape level involves the costly removal of combustible fuels. Despite significant progress, the number of acres in need of fuels reduction treatment is very large and not expected to be treated over the next decade [@thompson2023]. Wildfire suppression and response currently does and will continue to play an important role in protecting people, property, and other valued assets from wildfire. As federal firefighting agencies seek to improve efficient use of resources, the procurement and use of aviation resources presents several opportunities.

The current structure of procuring aviation resources involves exclusive use and call-when-needed contracts. Federal agencies should strive to align their use of aviation with operational demands. However, risk aversion may lead agency administrators to over contract for aviation resources to avoid public and political consequences that arise after destructive wildfire events. The question of contract structure is related to the question of optimal composition of the aircraft fleet addressed in a 2012 report by the Rand Corporation [@keating2012].

@@

![Trends of Extended Attack (top) and Initial Attack (middle) fires from 2007 to 2021. The points represent the median and the lines represent the interquartile range (25th to 75th percentile). The bottom panel represents the number of Large Airtankers (LAT) and Very Large Airtankers (VLAT) available by day under exclusive use contract in 2022. The initial and extended attack fire counts are compiled from historical daily situation reports posted at https://www.nifc.gov/nicc-files/sitreprt.pdf and collected using code from [@nguyen2023], and the number of LATs and VLATs contracted are from the June 24, 2022 Schedule of Items posted at https://www.fs.usda.gov/Internet/FSE_DOCUMENTS/fseprd1041508.pdf](../figures/ia_ea_fires.png){#fig-aviation_use}
